---
title: "Asynchronous Programming in R"
subtitle: "To kick-Start Developing a Scalable `shiny` App"
author:
- name: Kyle Chung
  affiliation:
date: "`r format(Sys.time(), '%d %b %Y')` Last Updated (14 Sep 2019 First Uploaded)"
output:
  html_notebook:
    highlight: pygments
    number_sections: yes
    theme: paper
    toc: yes
    toc_depth: 4
    toc_float: yes
    includes:
      in_header: /tmp/meta_header.html
  code_download: true
bibliography: async_r.bib
abstract: |
  The notebook is a practical walk-through of implementing asynchronous programming in R using package `future` and `promises`, mainly for the purpose of kick-starting developing a scalable `shiny` application. Indeed, all 3 packages come with very well-structured official tutorials already. This notebook serves more as a minimally sufficient one-stop reference for developers to quickly hands-on on the topic and get ready for the actual application development.
---

```{r meta, include=FALSE}
meta_header_file <- file("/tmp/meta_header.html")

# Add open graph meta.
meta <- c(
  '<meta name="author" content="Kyle Chung">',
  '<meta property="og:title" content="Asynchronous Programming in R">',
  '<meta property="og:type" content="article">',
  '<meta property="og:url" content="https://everdark.github.io/k9/programming/async/async_r.nb.html">',
  '<meta property="og:image" content="https://everdark.github.io/k9/assets/avatar.jpg">',
  '<meta property="og:description" content="One-stop quick hands-on for developing scalable shiny application with asynchronous programming.">'
)
contents <- meta

# Add Github corner.
github_corner_svg <- "../../../assets/github_corner.html"
github_corner_conf <- list(github_link="https://github.com/everdark/k9/tree/master/programming/r/async")
contents <- c(contents, stringr::str_interp(readLines(github_corner_svg), github_corner_conf))
writeLines(contents, meta_header_file)

close(meta_header_file)
```

```{r setup, include=FALSE}
# Create a hook for code chunk execution time.
knitr::knit_hooks$set(timeit=local({
  now=NULL
  function(before, options) {
    if ( before ) {
      now <<- Sys.time()
    } else {
      d <- difftime(Sys.time(), now)
      now <<- NULL
      paste(sprintf("=== Execution Time: %s Sec ===", as.character(d)))
    }
  }})
)
```

# Using `future`

```{r import_future}
library(future)
print(installed.packages()["future", "Version"])
```

The package `future` (@future) allows a single-threaded R process to launch a task in another thread to achieve asynchronous programming.
It is the main entry point for a task to be asynchronous.

To make an expression a `Future`,
we simply wrap it with a `future` function call:

```{r future}
f <- future({
  # Supposedly anything expensive to compute here.
  1 + 1
})
```

Then we can collect the result of a `Future`:

```{r value}
print(value(f))
```

In order for a `Future` to be useful,
we need to configure the execution plan.
That is,
how we'd like to run a given task in a separate process.

## Execution Plan

### Sequential Plan

The default execution plan for `future` is `sequential`:

```{r future_plan}
plan()  # Check the current execution plan.
```

It means that all function calls are executed sequentially in the current R session.
So the following future expression will block the session for 1 second:^[In this notebook we use a custom `knit_hooks` to time the relevant code chunk to showcase the blocking/non-blocking execution time.]

```{r sequential_future, timeit=TRUE}
plan(sequential)  # Set explicitly to a sequential plan.

f <- future({
  Sys.sleep(1)
  "I'm from the future!"
})

str(f)  # A SequentialFuture object.
```

The return value from a `future` is refered to as a *promise* in the asynchronous programming terminology.
As we discussed already,
to access the bounded value from a `future` expression (i.e., a promise) we use `value` function:

```{r future_value}
print(value(f))
```

Or a syntactic sugar can be used:

```{r future_sugar, timeit=TRUE}
v %<-% {
  Sys.sleep(1)
  "I'm from the future!"
}

print(v)
```

### Multi-Processing Plan

A `SequentialFuture` is of course not very useful at all.
The advantage of asynchgronous programming is to have multi-processing execution plan such that our function calls become non-blocking (asynchronous).
To enable such setting:

```{r future_plan_multiprocess}
plan(multiprocess)
plan()
```

Depending on the running platform,
the actual execution plan will be either `multicore` using fork (for Linux and macOS) or `multisession` using multiple R sessions (for Windows).
So `multiprocess` is just a platform-independent convenience plan for multi-threading.

It is worth noting that the behavior is technically NOT the same between `multicore` and `multisession`,
especially when it comes to variable scoping,
which we will discussed latter.

Under a `multiprocess` plan now our `future` expression becomes non-blocking:

```{r multiprocess_future, timeit=TRUE}
f <- future({
  Sys.sleep(1)
  "I'm from the future!"
})  # The call will return immediately.

print(class(f))  # A MultiprocessFuture object.
```

Again we can use `value` to access the bounded value.
But keep in mind that *`value` is a blocking call.*
If the promised value is not yet `resolved`,
it will wait until `resolved`:

```{r value_blocking, timeit=TRUE}
f <- future({
  Sys.sleep(1)
  "I'm from the future!"
})  # The call will return immediately.

print(value(f))  # This is blocking since `value` is waiting for `f` to be resolved.
```

One interesting fact about the pipe sugar `%<-%` in a `multiprocess` plan is that it is still non-blocking even if it seems to contain the `value` call implicitly:

```{r future_sugar_nonblocking, timeit=TRUE}
v %<-% {
  Sys.sleep(1)
  "I'm from the future!"
}
```

This is because `%<-%` is *lazy* in its implicit `value` call.
Only when the value is actually used will the call to `value` be executed.^[Readers should not confuse the laziness here with the `lazy` function argument in a `future` call. By setting `lazy=FALSE` (which is the default) in a `future` call it enables the expression to start execution immediately, otherwise not.]

To see this in action:

```{r future_sugar_blocking, timeit=TRUE}
v %<-% {
  Sys.sleep(1)
  "I'm from the future!"
}  # Non-blocking.

print(v)  # Blocking.
```

When using the future pipe operator,
one can still access the promise object without `value`.
This is done by `futureOf`:

```{r future_sugar_get_future}
v %<-% {
  Sys.sleep(1)
  "I'm from the future!"
}  # Non-blocking.

print(class(futureOf(v)))  # A Future object.
```

### Nested Futures

`future` expression can be nested and the corresponding plan can be configured accordingly.
By default the second-layer `future` will be `sequential` no matter what plan the first layer uses.

To specify the so-called *future topology*,
for example a `sequential` for the first layer and a `multiprocess` for the second layer,
we can give a list to the `plan` call:

```r
plan(list(sequential, multiprocess))
```

## Scope of a `future`

The scope of a `future` expression can be tricky.
In this section we will spend some time experimenting a bit on its scoping nature.

### Global Variables

Since `future` relies on a separate R process to handle asynchronous calls,
how does it deal with global variable in the master process?
By default `future` will automatically scan the given expression and determine which variable to *copy* for the separate thread to access.
This is controlled by the following call:

```
future(expr, envir=parent.frame(), globals=TRUE)
```

Setting `globals=TRUE` enables the auto-search feature,
which in theory can result in error by overlooking variables required by `expr`.
The search includes the `envir` (by default it is the calling `environment`) AND its parents until found (a.k.a. *lexical scoping*):

```{r, future_global_search}
x <- 42
y <- 64  # Search up to here for y.
k <- function() {
  x <- 0  # Search up to here for x.
  g <- function() {
    f <- future({
      list(x=x, y=y)
    }, globals=TRUE)
    value(f)
  }
  g()
}
k()
```

The search behavior can be controlled explicitly by setting `globals` to a variable name vector and specify the `enviroment` to search for:

```{r future_global_explicit}
x <- 42  # Exposed to future explicitly.
y <- 64  # Exposed to future explicitly.
k <- function() {
  x <- 0  # This is ignored since it is not in the specified search environment.
  g <- function() {
    f <- future({
      list(x=x, y=y)
    }, globals=c("x", "y"), envir=.GlobalEnv)
    value(f)
  }
  g()
}
k()
```

### Inconsistency between Plans

When it comes to scoping,
`future` is not very consistent across different execution plan.
In this section we discuss several notable examples that can be confusing.

#### Auto-Search Scope Inconsistency {-}

To see how the global is NOT copied correctly by the auto-search feature,
consider the following example:

```{r future_global_not_found}
plan(multisession)
z <- 10
f <- future({
  get("z", envir=environment(), inherits=TRUE)  # Error.
})
tryCatch(value(f), error=function(e) print(e))
```

A contrasting example is to make the expression easy for the `future` to understand that we need the variable:

```{r future_global_found}
plan(multisession)
z <- 10
f <- future({
  get("z", envir=environment(), inherits=TRUE)  # This now is working even BEFORE the eval to z.
  z  # This makes the auto-search work and export the variable from global.
})
tryCatch(value(f), error=function(e) print(e))
```

But the same function call works perfectly under either `multicore` or `sequential`:

```{r future_global_found_sequential}
plan(sequential)
z <- 10
f <- future({
  get("z", envir=environment(), inherits=TRUE)
})
tryCatch(value(f), error=function(e) print(e))
```

```{r future_global_found_multicore}
plan(multicore)  # This is not supported in Windows and will fallback to a sequential plan.
z <- 10
f <- future({
  get("z", envir=environment(), inherits=TRUE)
})
tryCatch(value(f), error=function(e) print(e))
```

#### Global Copy Inconsistency {-}

Remember that all globals are *copied* to the `future` block,
which means in-place change will not reflect on the original object even if the object is mutable:

```{r future_global_copy_multisession}
plan(multisession)
e <- new.env()  # A global mutable.
e$x <- 0

f <- future({
  e$x <- 42  # `e` is automatically searched and accessable, but changed in a copy.
})
invisible(value(f))

ls.str(e)  # The original copy is intact.
```

```{r future_global_copy_multicore}
plan(multicore)  # This is not supported in Windows and will fallback to a sequential plan.
e <- new.env()  # A global mutable.
e$x <- 0

f <- future({
  e$x <- 42
})
invisible(value(f))

ls.str(e)
```

However,
the above fact only holds for a `multiprocess` plan.
Under `sequential` plan the global original is indeed changed in-place:

```{r future_global_copy_sequential}
plan(sequential)
e <- new.env()  # A global mutable.
e$x <- 0

f <- future({
  e$x <- 42
})
invisible(value(f))

ls.str(e)  # The original copy has been modified!
```

#### Super Assignment Inconsistency {-}

Super assignment (`<<-`) behaves differently by execution plan.
For a `sequential` plan:

```{r future_scope_superassign_sequential}
plan(sequential)
x <- 0
f <- future({
  x <<- 42
  y <- 0
})
invisible(value(f))
x
```

But for a `multisession` plan:

```{r future_scope_superassign_multisession}
plan(multisession)
x <- 0
f <- future({
  x <<- 42
  y <- 0
})
invisible(value(f))
x
```

Or for a `multicore` plan:

```{r future_scope_superassign_multicore}
plan(multicore)  # This is not supported in Windows and will fallback to a sequential plan.
x <- 0
f <- future({
  x <<- 42
  y <- 0
})
invisible(value(f))
x
```

## Non-Blocking Resolution Check

Since now a `future` expression may be or may not be `resolved` already,
and a `value` call to the promised value will block if the value is not yet `resolved`,
it becomes important to have the ability that allows us to check whether a promise is already `resolved`,
without being blocked.

This is exactly what the function `resolved` is doing:

```{r future_resolved}
plan(multiprocess)

f <- future({
  Sys.sleep(1)
  "I'm from the future!"
})

resolved(f)  # Return immediately and `f` is not yet resolved.

Sys.sleep(1)

resolved(f)  # Return immediately and `f` should be already resolved.
```

Now it is theoretically possible to create one non-blocking future thread to wait for another non-blocking future thread:

```{r nonblocking_future_wait_for_future}
# Educational purpose only.
# This pattern may not be very useful in practice.
f1 <- future({
  Sys.sleep(3)
  "I'm from the future!"
})

f2 <- future({
  while ( TRUE ) {
    if ( resolved(f) ) {
      value(f)
      break
    }
  }
})
```

Whether such pattern is useful at all depends on the actual use case.
But in [the next section](#promises) we will learn how to control even more on a promise to arrive at a full-fletched asynchronous programming framework in R.

## Error Propogation

Error from a `future` expression will propogate to the actual bounded value but not the `Future` object itself.

```{r future_error}
fe <- future(stop("Error from the future!"))
print(class(fe))  # No exception before access the value.

tryCatch(value(fe), error=function(e) print(e))
```

For explicit error handling it is better to resort to using `promises`.

# Using `promises` {#promises}

Using `promises` (@promises) is one big step ahead of `future`,
enabling even more flexibility on asynchronous programming.
But it also drastically changes how we should write our code--specifically, in a promise-style.

```{r import_pormises}
library(promises)
print(installed.packages()["promises", "Version"])
```

```{r multiprocess_plan_promises}
# Fix the plan to multiprocess for promises.
plan(multiprocess)
```

## From `future` to `promises`

A `future` expression return a *promise*.
Such promise can be converted explicitly to a `promise` object using `then`:

```{r promise}
f <- future({
  Sys.sleep(1)
  "I'm from the future!"
})

p <- then(f, onFulfilled=function(v) v)
print(class(p))  # A promise object.
str(p)
```

Unlike the original return value from a `future`,
a promise,
even under `multiprocess`,
is always `resolved`:

```{r always_resolved_promise, timeit=TRUE}
f <- future({
  Sys.sleep(1)
  "I'm from the future!"
})

p <- then(f, onFulfilled=function(v) v)
resolved(p)  # This is always, immediately, TRUE.
```

Put it differently,
`resolved` only works (or is only meaningful) for a `Future` object but not for a `promise` object.

We loosely call the return value of a `future` as a promise as well since this is the common wording used in asynchronous programming.
But here we explicitly refer to the object class `promise` implemented by the R package `promises`.
When we refer to the general concept of promise,
we will avoid using the syntax highlight for code for readers' ease of distinguishing the difference.

A `Future` is not,
but can be a `promise`.
Indeed,
when a `Future` object is fed to a `then` function,
it will be immediately converted to a `promise`-like object by attribute assignment.
This can be easily seen in the following code:

```{r future_implicit_promise}
f <- future({
  Sys.sleep(1)
  "I'm from the future!"
})
str(f)  # A Future, before called with a then function.

p <- then(f, onFulfilled=function(v) v)
str(f)  # The same Future, after called with then and has been attached promise attributes.
```

A `Future` is either `resolved` or not `resolved`.
While a `promise` is either *pending*, *fulfilled*, or *rejected* (due to error).

```{r promise_status}
f <- future({
  Sys.sleep(.1)
  "I'm from the future!"
})

p <- then(f, onFulfilled=function(v) v)
print(p)  # Still pending.
```

Due to the specialty of the notebook rendering environment,
a promise is always shown as pending no matter how long we wait for.
For an actual R session the result will be something like:

```
<Promise [fulfilled: character]>
```

## No Way Back Once `promise`d

How do we extract the bounded value from a `promise`,
like what we do with a `Future` by using the `value` function?

It turns out that,
**we cannot.**

Indeed,
**a `promise` is ALWAYS a `promise`.**
There is simply no way back once we pipe our task into a `promise`.
The design philosophy is that we never know when the `promise`d value will be available,
and hence the return value from a `promise` must always be a `promise`.

To process the return value from a `promise`,
we simply chain it with another `promise`:

```{r promise_chain}
f <- future(1 + 1)
p1 <- then(f, onFulfilled=function(v) v + 1)
p2 <- then(p1, onFulfilled=function(v) v + 2)
```

## Error Handling

When a `then` job failed at either its `onFulfilled` or `onRejected` task,
it returns a rejected `promise` with the corresponding error type.
Rejection does NOT propagate, though.
When a `then` job processes a rejected `promise`,
it will go to the `onRejected` branch (by default simply propagate the error if not specified) and if that task is done without error,
the resulting value is a fulfilled `promise`.

The following example illustrates the above idea:

```{r promise_error_1}
plan(sequential)

f <- future(1 + 1)

# The first promise will fail at its onFuilfilled task (on purpose).
p1 <- then(
  f,  # This is a fulfilled promise.
  onFulfilled=function(v) stop(v),  # Hence we follow this branch. (And then will fail.)
  onRejected=function(e) e
)

# The second promise will go to the onRejected branch since the first promise failed.
p2 <- then(
  p1,  # This is failed on its own onFulfilled task.
  onFulfilled=function(v) print(sprintf("From onFulfilled: %s", v)),
  onRejected=function(e) print(sprintf("From onRejected: %s", e))  # Hence we go here.
)



p1
p2
```

The first `promise` object `p1` will have a value of

```
<Promise [rejected: simpleError]>
```

after running its own `onFulfilled` task.
But the second `promise` object `p2` instead will have a value of

```
<Promise [fulfilled: character]>
```

after running its own `onRejected` task and also print the message

```
[1] "From onRejected: Error in onFulfilled(value): 2"
```

Then same logic applies to `Future` object (i.e., converted `promise`):

```{r promise_error_2, eval=FALSE}
# Error at the future expression.
f <- future(stop("Something just went wrong at the very beginning."))

# The first promise will go for the onRejected task, and return successfully.
p1 <- then(
  f,
  onFulfilled=function(v) {
    print("From onFulfilled.")
    v
  },
  onRejected=function(e) {
    print("From onRejected")
    e
  }
)

# Since the onRejected task is successfully excecuted, p1 becomes a fulfilled promise.
# p2 then will go for the onFulfilled task.
p2 <- then(
  p1,
  onFulfilled=function(v) {
    print("From onFulfilled.")
    v
  },
  onRejected=function(e) {
    print("From onRejected")
    e
  }
)

Sys.sleep(3)  # Wait longer to allow for potential backend overhead.

p1
p2
```

This time `p1` will print

```
[1] "From onRejected"
```

with a value of

```
<Promise [fulfilled: simpleError]>
```

And `p2` will print

```
[1] "From onFulfilled."
```

with a value of

```
<Promise [fulfilled: simpleError]>
```

Both `promise`s are `fulfilled` (i.e., no further error at the branch) on their corresponding task,
whether it is a branch task of `onFulfilled` or `onRejected`.

## Syntactic Sugars

There are several syntactic sugars available when using the `then` API.

### Lambda Expression {-}

```{r promise_sugar_lambda}
f <- future({
  Sys.sleep(.1)
  "I'm from the future!"
})

p <- then(f, ~{
  .  # Return the Future value as-is.
})
```

### Promise-Aware Pipe {-}

```{r promise_sugar_pipe}
f <- future({
  Sys.sleep(.1)
  "I'm from the future!"
}) %...>% {
  cat(.)  # This won't output to the notebook code chunk.
}
```

The pipe `%...>%` only supports `onFulfilled` function.
In such case the `onRejected` task is simply an error propagation.

For `onRejected` function one can use `%...!%` instead.
And the `onFulfilled` task is a simple `identity`.

For a complete custom branch handling one should always use the `then` API explicitly.

# Async with Reactive Programming

Reactive programming in R is introduced by the well-known `shiny` package (@shiny) which facilitates the ease of web application development purely using R,
usually for data-driven dashboard building purpose.

A huge limitation about such web app is that R is a single-threaded process.
Reactive programming itself does not provide asynchrony.
In order for a web app to be scalable for a multi-session use case,
reactive programming must combine with asynchronous programming.

The good news is that `shiny` has come fully support for `future` and `promises` under its reactive programming framework after its major release of v1.1.

## A Quick Recap: Reactive Programming

```{r import_shiny}
library(shiny)
print(installed.packages()["shiny", "Version"])
```

Broadly speaking there are two types of handler in reactive programmning for `shiny`:
value handler and event handler.

### Value Handler

To register an object whose value changes reactively (according to, say, user input from a web app portal),
we can use `reactiveVal` or `reactiveValues`.
`reactiveVal` is designed for a single value (single object) while `reactiveValues` is designed for a list of values (multiple objects).

To initialize a reactive value without a default (`NULL` as default):

```r
r <- reactiveVal(value=NULL)
```

To update the value we simply call:

```r
r(x)  # `x` is the new value assigned reactively to `r`.
```

And to retrieve the value we can call without argument:

```r
r()
```

In general `reactiveVal` is only callable under a reactive context (all the `render*` function such as `renderText`, `renderPrint`).
To test it interactively,
we can also use `isolate` to directly retrieve the value non-reactively:

```{r reactiveVal_isolate}
r <- reactiveVal(value=NULL)
isolate(r())
r(42)
isolate(r())
```

To initialize a list of reactive values:

```r
rl <- reactiveValues()
```

`reactiveValues` can be directly update by using the list assignment syntax:

```r
rl$a <- 1
rl$b <- 1
```

To test it:

```{r reactiveValues_isolate}
rl <- reactiveValues()
rl$a <- 1
rl$b <- 1
isolate(rl$a + rl$b)
```

We can create reactive functions which depend on reactive values.
The function will re-execute everytime any of the dependent reactive values change.

```{r reactive_function}
rl <- reactiveValues(a=1, b=2)

rf <- reactive({
  rl$a + rl$b
})

isolate(rf())
```

Or we can create reactive observer functions similar to `reactive` but only for its side-effects (i.e., no return value):

```{r reactive_observer}
rl <- reactiveValues(a=1, b=2)

# Nothing will be printed since the value doesn't change without a reactive context.
observe({
  print(rl$a + rl$b)
})
```

### Event Handler

We can also create handlers that explicitly respond to UI component or other reactive events defined by a reactive function.

For example,
we can react to a user button click by a `eventReactive` function:

```r
r <- eventReactive(input$run, {expr})  # Assume `run` is the id for a button in the UI code.
```

Not just UI component,
`eventReactive` can react to anything reactive:

```{r eventReactive}
r <- reactiveVal()
r(123)
er <- eventReactive(r(), 456)
isolate(er())
```

We can do the same but only for side-effects by using `observeEvent`:

```r
observeEvent(input$run, {expr})  # Assume `run` is the id for a button in the UI code.
```

### A Minimum App Template

For completeness,
the following code chunk provides a minimum single-file working `shiny` app with a user input button and a output text window.

```{r min_app, eval=FALSE}
# Define frontend code.
ui <- fluidPage(

  titlePanel("Test Reactive Programming in R"),
  actionButton("do", "Do something."),
  uiOutput("out")

)

# Define backend code.
server <- function(input, output, session) {

  observeEvent(input$do, {
    output$out <- renderText({
      "Something."
    })
  })

}

# Launch the app.
shinyApp(ui=ui, server=server, options=list(port=8787))
```

## Promise-Aware Reactive Expression

All reactive functions provided by `shiny` are both `future` and `promise` aware.
It means that reactive expressions can be a `Future`:

```{r reactive_future}
rl <- reactiveValues(a=1, b=2)

rf <- reactive({
  future(rl$a + rl$b)
})

isolate(value(rf()))
```

or a `promise`:

```{r reactive_promise}
rl <- reactiveValues(a=1, b=2)

rf <- reactive({
  future(rl$a + rl$b) %...>% {
    .
  }
})

isolate(str(rf()))
```

## Experimental App on Async

To see how one session is blocking another in a single-threaded app,
here is a minimum app for pure illustration purpose:

```{r shiny_async_app, eval=FALSE}
library(shiny)
library(future)
library(promises)


exec_plan <- commandArgs(trailingOnly=TRUE)[1]
if ( is.na(exec_plan) ) exec_plan <- "sequential"

plan(exec_plan)


# Define frontend code.
ui <- fluidPage(

  titlePanel("Async Shiny App"),
  textOutput("time"),
  actionButton("do", "Do some heavy works."),
  verbatimTextOutput("out")

)


do_heavy_work <- function() {
  st <- Sys.time()

  Sys.sleep(5) # Or anything expensive here.

  et <- Sys.time()
  list(st=st, et=et)
}


# Define backend code.
server <- function(input, output, session) {

  output$time <- renderText({
    invalidateLater(1000, session)
    paste("The current time is", Sys.time())
  })

  observeEvent(input$do, {
    st <- Sys.time()  # This only record when the app starts process the input but NOT when the user hit the button.
    output$out <- renderText({
      future(do_heavy_work()) %...>% {
        paste(
          "Heavy work done!",
          sprintf("Started at %s", st),
          sprintf("Ended at %s", .$et),
          sprintf("Time used: %s", .$et - st),
          sep="\n"
        )
      }
    })
  })

}

# Launch the app.
shinyApp(ui=ui, server=server, options=list(port=8787))
```

To play around with the actual app,
run:

```sh
Rscript app.R
```

for synchronous mode.

The app will be listening on `127.0.0.1:8787`.
Open more than 1 tab and click the `do` button for all the opening tabs.
For a `sequential` app we shall observe the timer stop jumping when there is any other session (browser tab) still working and hence blocking.
The more other sessions are still working,
the more times you will see the timer jump and stop.
The start time recorded in the output text is always roughly 5 seconds before it ended.
However the start time is NOT the time we hit the `do` button.

Now close the app.
For asynchronous mode run:

```sh
Rscript app.R multiprocess
```

and investigate again.

The timer will still stop jumping when we hit the `do` button.
But this time it didn't stop more than once.
*It only stop for its own session.*

And the timer is stopping because we didn't implement the timer asynchronously.
In the experimental app only the `do_heavy_work` operation is implemented in asynchronous mode.

In general we will only make those heavy tasks asynchronous instead of trying to make everything asynchronous.

# Closing Remarks

Up to now we've equiped with the basic knowledge to start developing asynchronous web application using R.
There are far more features then what we just discussed here for all these packages:
`future`, `promises`, and `shiny`.
But we will leave it for exploration and retain the scope of this notebook at a entry-level.

# References
