---
title: "Visualization for TW General Election"
subtitle: "2020 Presidential and Legislative Results"
author:
- name: Kyle Chung
  affiliation:
date: "`r format(Sys.time(), '%d %b %Y')` Last Updated (19 Jan 2020 First Uploaded)"
output:
  html_notebook:
    code_folding: hide
    fig_caption: yes
    fig_width: 8.33
    fig_height: 6.25
    highlight: tango
    includes:
      in_header: /tmp/meta_header.html
    theme: paper
    toc: yes
    toc_float: yes
  code_download: true
bibliography: tw_election_2020.bib
nocite: |
  @data.table
  @ggplot2
  @ggrepel
  @plotly
  @showtext
  @sf
  @lmtest
  @sandwich
link-citations: yes
abstract: |
  The notebook provides a visual description on the 2020 Taiwanese presidential and legislative election results. The polling took place at 2020-01-11 and data has been collected by Central Election Commision (CEC) at the polling place level.
---

```{r meta, include=FALSE}
meta_header_file <- file("/tmp/meta_header.html")

# Add open graph meta.
meta <- c(
  '<meta name="author" content="Kyle Chung">',
  '<meta property="og:title" content="2020 Taiwanese Presidential Election Data">',
  '<meta property="og:type" content="article">',
  '<meta property="og:url" content="https://everdark.github.io/k9/projects/tw_election_2020/tw_election_2020.nb.html">',
  '<meta property="og:image" content="https://everdark.github.io/k9/assets/tw_election_2020.png">',
  '<meta property="og:description" content="A data science notebook about 2020 Taiwanese presidential election.">'
)
contents <- meta

# Add Github corner.
github_corner_svg <- "../../assets/github_corner.html"
github_corner_conf <- list(github_link="https://github.com/everdark/k9/tree/master/projects/tw_election_2020")
contents <- c(contents, stringr::str_interp(readLines(github_corner_svg), github_corner_conf))
writeLines(contents, meta_header_file)

close(meta_header_file)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align="center")
```

# Prerequisites {.tabset .tabset-fade .tabset-pills}

## Platform

The source data contains traditional Chinese characters encoded in UTF-8.
To avoid troublesome encoding issues (especially in a Windows platform with a default `CP1252` code page),
the **codes in this notebook are only tested on Ubuntu and macOS.**
Codes are in general still working under Windows.
Just that some code chunks may not be able to print out unicode characters nicely.

## Data

### Election Data {-}

We use minimally pre-processed data for our analytics.
Please refer to the repo [TW_Presidential_Election_2020](https://github.com/everdark/TW_Presidential_Election_2020) for details about the pre-processing logic.
To directly download the archives please visit [here](https://github.com/everdark/TW_Presidential_Election_2020/releases/latest).
We put the uncompressed files under `data/processed` at our project root directory.

### Map Data {-}

For Taiwan map information,
the `shp` file ([shapefile](https://en.wikipedia.org/wiki/Shapefile)) can be downloaded at [GADM.org](https://gadm.org/download_country_v3.html).
We put the unzipped files under `data/map` at our project root directory.

## Dependencies

Additional dependencies are required for Linux machine in order to do map plotting.
Here is the instruction for Ubuntu.

Install [`gdal`](https://gdal.org/):

```sh
sudo add-apt-repository ppa:ubuntugis/ubuntugis-unstable
sudo apt update
sudo apt install libgdal-dev
```

Install `ubunits`:

```sh
sudo apt install libudunits2-dev
```

After installation of the above packages,
we can then install and import the following R packages for this project:

```{r import, results="hide", message=FALSE, warning=FALSE, class.source="fold-show"}
library(data.table)

# For viualization.
library(ggplot2)
library(ggrepel)
library(plotly)

# Fix macOS ggplot device to locate a Chinese font.
# To make `knitr` aware of this we also need to speficy `fig.showtext=TRUE` for chunks with ggplot output.
library(showtext)
font_add_google("Noto Serif TC", "noto")
showtext_auto()

# For spatial data visualization.
library(sf)

# For linear modeling with robust std err.
library(lmtest)
library(sandwich)
```

# Background

+ [2020 10th Taiwanese Presidential Election](https://en.wikipedia.org/wiki/2020_Taiwanese_presidential_election)
+ [2020 15th Taiwanese Legislative Election](https://en.wikipedia.org/wiki/2020_Taiwanese_legislative_election)

# Visual Analytics on Election Results {.tabset .tabset-fade .tabset-pills #head}

## 第十五屆總統選舉結果 The 15th Presidential Election Results {.tabset .tabset-fade .tabset-pills}

### 有效票 Valid Vote Counts {.tabset .tabset-fade .tabset-pills}

#### By County

```{r parse_tw_county_map}
# Read shapefile for spatial information about counties in Taiwan.
tw_county_sf <- sf::st_read("data/map/gadm36_TWN_2.shp", quiet=TRUE)

# Align county naming.
tw_county_sf$NL_NAME_2 <- gsub("台中", "台中市", tw_county_sf$NL_NAME_2)
tw_county_sf$NL_NAME_2 <- gsub("台南", "台南市", tw_county_sf$NL_NAME_2)
tw_county_sf$NL_NAME_2 <- gsub("馬祖列島", "連江縣", tw_county_sf$NL_NAME_2)
```

```{r process_presidential_county, results="hide"}
# process county-level presidential election results.
p_county <- fread("data/processed/presidential/presidential_counties.csv", encoding="UTF-8")
p_county <- p_county[, Region:=trimws(Region, whitespace="[\\h\\v]")]  # Trim full-width white space.
p_county <- p_county[, Region:=gsub("臺", "台", Region)]  # Align character with shapefile.
setnames(p_county, "投票率H H=C÷G", "Turnout")

# Final counts.
cols_candidate <- c("(1) 宋楚瑜 余湘", "(2) 韓國瑜 張善政", "(3) 蔡英文 賴清德")
p_final <- as.data.table(colSums(p_county[, cols_candidate, with=FALSE]), keep.rownames=TRUE)
setnames(p_final, c("Candidate", "Votes"))
p_final <- p_final[, Share:=Votes / sum(Votes)]

# Merge with shp to get English county name.
p_county <- merge(p_county, tw_county_sf[, c("NL_NAME_2", "NAME_2")],
                  by.y="NL_NAME_2", by.x="Region")
p_county$geometry <-NULL
setnames(p_county, "NAME_2", "RegionEn")
p_county <- p_county[, County:=paste(Region, RegionEn)]

# Melt from wide to long for plotting purpose.
cols_region <- c("County", "Region", "RegionEn")
p_county_l <- melt(p_county[, c(cols_region, cols_candidate), with=FALSE],
                   id.vars=cols_region, variable.name="Candidate", value.name="Votes")
p_county_l <- p_county_l[, Candidate:=factor(Candidate)]  # Level by candidate number.

# Compute difference between DPP and KMT by county.
setorder(p_county_l, County, Candidate)
p_county_l <- p_county_l[, Votes_Diff:=Votes[3] - Votes[2], by=.(County)]

# Compute total votes by county.
p_county_l <- p_county_l[, Total_Votes:=sum(Votes), by=.(County)]

# Compute share by county.
p_county_l <- p_county_l[, Votes_Share:=Votes / Total_Votes]
p_county_l <- p_county_l[, Share_Diff:=Votes_Share[3] - Votes_Share[2], by=.(County)]
p_county_diff <- unique(p_county_l, by="County")[
  , c(cols_region, "Total_Votes", "Share_Diff"), with=FALSE]
p_county_diff <- p_county_diff[, dpp_win:=Share_Diff > 0]

# Join share diff to spatial dataframe.
tw_county_sf <- merge(tw_county_sf, p_county_diff, by.x="NL_NAME_2", by.y="Region")

# Compute difference between DPP and KMT on the wide data as well.
p_county <- p_county[, Vote_Diff:=get("(3) 蔡英文 賴清德") - get("(2) 韓國瑜 張善政")]
p_county <- p_county[, Share_Diff:=Vote_Diff / get("有效票數A A=1+2+...+N")]

# Color configuration for main parties.
main_party_colors <- c("darkorange", "dodgerblue", "green3")
```

```{r presidential_final_result, fig.showtext=TRUE, fig.cap="2020 Presidential Election Final Outcome"}
ggplot(p_final, aes(x=Candidate, y=Votes, fill=Candidate)) +
  geom_bar(stat="identity") +
  labs(x="候選人", y="得票數",
       title="2020台灣總統大選結果") +
  geom_text(aes(x=Candidate, y=Votes / 2,
                label=sprintf("%s\n(%s)", scales::comma(Votes), scales::percent(Share))),
            size=5) +
  scale_fill_manual(values=main_party_colors) +
  scale_y_continuous(labels=scales::comma) +
  theme(legend.position="none") +
  coord_flip()
```

The pre-processed county result for presidential election is a *wide* format by county:

```{r show_presidential_county_result}
head(p_county[, -c(1:3)])
```

And we've reshaped it into a *long* format for ease of plotting:

```{r show_presidential_taiepi}
# Show result for Taipei.
p_county_l[Region == "台北市"]
```

Plot vote counts by county:

```{r presidential_county_bar, fig.cap="2020 Presidential Election Vote Counts by County"}
p <- ggplot(p_county_l, aes(x=reorder(County, Total_Votes), y=Votes, fill=Candidate)) +
  geom_bar(stat="identity", position="dodge", width=.8) +
  scale_fill_manual(name="候選人", values=main_party_colors) +
  scale_y_continuous(labels=scales::comma) +
  labs(x="縣市（依總票數排序）", y="得票數",
       title="2020台灣總統大選候選人各縣市：得票數") +
  coord_flip() +
  theme(legend.position="bottom")

ggplotly(p, width=800, height=600)
```

And also vote shares (vote counts / total valid votes):

```{r presidential_county_bar_pct, fig.cap="2020 Presidential Election Vote Shares by County"}
p <- ggplot(p_county_l, aes(x=reorder(County, Total_Votes), y=Votes_Share, fill=Candidate)) +
  geom_bar(stat="identity", position="stack", width=.8) +
  geom_text(aes(label=scales::percent(Votes_Share)), position=position_stack(vjust=.5), size=2) +
  scale_fill_manual(name="候選人", values=main_party_colors) +
  labs(x="縣市（依總票數排序）", y="得票率",
       title="2020台灣總統大選候選人各縣市：得票率") +
  coord_flip() +
  theme(legend.position="bottom")

ggplotly(p, width=800, height=600)
```

Since the main parties are DDP and KMT,
we define vote share difference as vote share of DDP minus that of KMT.
The following barplot gives us a clear view of which county DPP won the most.

```{r presidential_county_bar_pct_diff, fig.cap="2020 Presidential Election Vote Share Difference (DPP's - KMT's) by County"}
p <- ggplot(p_county_diff, aes(x=reorder(County, Share_Diff), fill=dpp_win, y=Share_Diff)) +
  geom_bar(stat="identity", width=.8, color="black", alpha=.5) +
  scale_fill_manual(values=c("dodgerblue", "green3")) +
  theme(legend.position="none") +
  geom_text(aes(label=scales::percent(Share_Diff)), position=position_stack(vjust=.5), size=2.5) +
  labs(x="縣市（依得票率差排序）", y="綠藍得票率差",
       title="2020台灣總統大選各縣市：綠藍得票率差") +
  coord_flip()

ggplotly(p, width=800, height=600)
```

We can also visualize the above ranking with a choropleth map:

```{r presidential_county_map, fig.cap="Choropleth Map for Vote Share Differences (DPP's - KMT's)"}
p <- ggplot(tw_county_sf, aes(text=County)) +
  geom_sf(aes(fill=Share_Diff)) +
  coord_sf(xlim=c(118, 122.5), ylim=c(21.5, 26.5)) +
  scale_fill_gradient(name="綠藍得票率差", low="blue3", high="green3") +
  labs(title="2020台灣總統大選各縣市：綠藍得票率差")

ggplotly(p, width=800, height=600)
```

Apparently DPP won most counties besides maintain and outlying island areas.

[Go Back to Top](#head)

#### By County Sub-Region {.tabset .tabset-fade .tabset-pills}

```{r process_presidential_region, results="hide"}
# Read county subregion-level presidential election results.
p_region <- fread("data/processed/presidential/presidential_regions.csv", encoding="UTF-8")
p_region <- p_region[, Region:=trimws(Region, whitespace="[\\h\\v]")]  # Trim full-width white space.
setnames(p_region, "投票率H H=C÷G", "Turnout")

# Compute difference between DPP and KMT by region.
p_region <- p_region[, Vote_Diff:=get("(3) 蔡英文 賴清德") - get("(2) 韓國瑜 張善政")]
p_region <- p_region[, Share_Diff:=Vote_Diff / get("有效票數A A=1+2+...+N")]

# Wide to long.
p_region_l <- melt(p_region[, c("By", "Region", cols_candidate), with=FALSE],
                   id.vars=c("By", "Region"), variable.name="Candidate", value.name="Votes")
p_region_l <- p_region_l[, Candidate:=factor(Candidate)]  # Level by candidate number.
p_region_l <- p_region_l[p_region, Share_Diff:=i.Share_Diff, on=.(Region)]

setorder(p_region, -Share_Diff)

# Group counties into areas.
north_counties <- c("臺北市", "新北市", "基隆市", "新竹市", "桃園市", "新竹縣", "宜蘭縣")
central_counties <- c("臺中市", "苗栗縣", "彰化縣", "南投縣", "雲林縣")
south_counties <- c("高雄市", "臺南市", "嘉義市", "嘉義縣", "屏東縣", "澎湖縣")
east_counties <- c("花蓮縣", "臺東縣")
island_counties <- c("金門縣", "連江縣")

# Helper function for share difference bar plot.
plot_region_share_diff <- function(counties, title) {
  DT <- p_region[By %in% counties]
  DT[, By:=factor(By, levels=counties)]  # Fix order for facet grids.
  DT[, Region2:=paste(By, Region)]  # Fixed order within facet (same-name region for different counties).
  DT[, dpp_win:=Share_Diff > 0]
  p <- ggplot(DT, aes(x=reorder(Region2, Share_Diff), y=Share_Diff, fill=dpp_win)) +
    geom_bar(stat="identity", width=.8, color="black", alpha=.5) +
    scale_fill_manual(values=c("dodgerblue", "green3")) +
    theme(legend.position="none") +
    geom_text(aes(label=scales::percent(Share_Diff)), position=position_stack(vjust=.5), size=2.5) +
    labs(x="", y="", title=title) +
    coord_flip() +
    facet_grid(By ~ ., scales="free_y", space="free_y")
  p
}
```

Vote share difference (DPP v.s. KMT) ranked by region across the nation:

```{r share_diff_rank_by_region}
p_region_rank <- p_region[, .(By, Region, Share_Diff, Vote_Diff)]
p_region_rank <- p_region_rank[, Share_Diff:=scales::percent(Share_Diff, .001)]
p_region_rank <- p_region_rank[, i:=.I]
setcolorder(p_region_rank, "i")
setnames(p_region_rank, c("排序", "縣市", "鄉鎮市區", "綠藍得票率差", "綠藍得票數差"))
p_region_rank[]
```

We bar-plot the above ranking by region by county:

##### North Taiwan

```{r presidential_north_region_bar_pct_diff, fig.showtext=TRUE, fig.height=17, fig.width=8}
plot_region_share_diff(
  north_counties,
  title="2020台灣總統大選北部地區（鄉鎮市區）：綠藍得票率差")
```

##### Central Taiwan

```{r presidential_central_region_bar_pct_diff, fig.showtext=TRUE, fig.height=20, fig.width=8}
plot_region_share_diff(
  central_counties,
  title="2020台灣總統大選中部地區（鄉鎮市區）：綠藍得票率差")
```

##### South Taiwan

```{r presidential_south_region_bar_pct_diff, fig.showtext=TRUE, fig.height=22, fig.width=8}
plot_region_share_diff(
  south_counties,
  title="2020台灣總統大選南部地區（鄉鎮市區）：綠藍得票率差")
```

##### East Taiwan

```{r presidential_east_region_bar_pct_diff, fig.showtext=TRUE, fig.height=6, fig.width=8}
plot_region_share_diff(
  east_counties,
  title="2020台灣總統大選東部地區（鄉鎮市區）：綠藍得票率差")
```

##### Outlying Islands

```{r presidential_island_region_bar_pct_diff, fig.showtext=TRUE, fig.height=3, fig.width=8}
plot_region_share_diff(
  island_counties,
  title="2020台灣總統大選離島地區（鄉鎮市區）：綠藍得票率差")
```

### 無效票 Invalid Vote Counts {.tabset .tabset-fade .tabset-pills}

[Invalid votes](https://en.wikipedia.org/wiki/Spoilt_vote) are votes cast but declared invalid.
There is no way to objectively record the intention of an invalid vote,
here we simply statistically describe the share of invalid votes from total votes cast.

#### By County

```{r process_presidential_invalid_votes_county, results="hide"}
# Process invalid votes by county.
p_county <- p_county[, Invalid_Votes_Share:=get("無效票數B") / get("投票數C C=A+B")]
p_county <- p_county[, Invalid_Votes_ShareP:=Invalid_Votes_Share * 100]

# Join invalid vote shares to spatial dataframe.
tw_county_sf <- merge(tw_county_sf, p_county[, .(Region, Invalid_Votes_ShareP)], by.x="NL_NAME_2", by.y="Region")
```

```{r presidential_county_bar_pct_invalid, fig.cap="2020 Presidential Election Invalid Vote Shares by County"}
p <- ggplot(p_county, aes(x=reorder(County, Invalid_Votes_Share), y=Invalid_Votes_ShareP)) +
  geom_bar(stat="identity", width=.8, fill="white", color="black") +
  geom_text(aes(label=scales::percent(Invalid_Votes_Share)), position=position_stack(vjust=.5), size=2.5) +
  labs(x="縣市（依廢票率排序）", y="廢票率％",
       title="2020台灣總統大選各縣市：廢票率") +
  coord_flip()

ggplotly(p, width=800, height=600)
```

To see the geolocational differences:

```{r presidential_county_map_invalid, fig.cap="Choropleth Map for Invalid Vote Shares"}
p <- ggplot(tw_county_sf, aes(text=County)) +
  geom_sf(aes(fill=Invalid_Votes_ShareP)) +
  coord_sf(xlim=c(118, 122.5), ylim=c(21.5, 26.5)) +
  scale_fill_gradient(name="廢票率％", low="white", high="black") +
  labs(title="2020台灣總統大選各縣市：廢票率％")

ggplotly(p, width=800, height=600)
```

Check if there is any correlation between invalid votes and the DPP-KMT differences:

```{r presidential_invalid_vsd_county, fig.showtext=TRUE, fig.cap="2020 Presidential Election Invalid Vote Shares vs DPP-KMT Share Differences by County"}
ggplot(p_county, aes(x=Invalid_Votes_Share, y=Share_Diff, label=County)) +
  geom_point(shape=4, size=4, stroke=2) +
  labs(x="廢票率", y="綠藍得票率差",
       title="2020台灣總統大選各縣市：廢票率Ｘ綠藍得票率差") +
  geom_text_repel(color="grey40")
```

[Go Back to Top](#head)

#### By County Sub-Region

```{r process_presidential_invalid_votes_region, results="hide"}
# Process invalid votes by region.
p_region <- p_region[, Invalid_Votes_Share:=get("無效票數B") / get("投票數C C=A+B")]
p_region <- p_region[, Invalid_Votes_ShareP:=Invalid_Votes_Share * 100]
```

```{r presidential_invalid_vsd_region, fig.cap="2020 Presidential Election Invalid Vote Shares vs DPP-KMT Share Differences by Region"}
p <- ggplot(p_region, aes(x=Invalid_Votes_Share, y=Share_Diff, label=paste(By, Region))) +
  geom_point(shape=4, size=4) +
  labs(x="廢票率", y="綠藍得票率差",
       title="2020台灣總統大選各鄉鎮市區：廢票率Ｘ綠藍得票率差") +
       geom_smooth(method="lm", color="black")
ggplotly(p, width=800, height=600)
```

A simple linear model on invalid vote shares against DPP-KMT share differences suggests the correlation is at least statistically significant.
But such correlation may be [*spurious*](https://en.wikipedia.org/wiki/Spurious_relationship) so the interpretation shall be cautious.

```{r lm_invalid}
lm_invalid <- lm(Share_Diff ~ Invalid_Votes_ShareP, data=p_region)
coeftest(lm_invalid, vcov=vcovHC(lm_invalid, type="HC0"))
```

[Go Back to Top](#head)

### 投票率 Voter Turnout {.tabset .tabset-fade .tabset-pills}

[Voter turnout](https://en.wikipedia.org/wiki/Voter_turnout) is defined as total votes cast (valid and invalid) divided by legitimate voters.
The information is readily available in the source data from the column `投票率H H=C÷G`.
In this section we explore the possible relation between turnout and vote counts.

#### By County

```{r process_presidential_county_turnout, results="hide"}
# Get voting rate from wide data to long.
p_county_l <- p_county_l[p_county, Turnout:=i.Turnout, on=.(County)]
```

```{r presidential_county_turnout_bar, fig.cap="2020 Presidential Election Voter Turnout by County"}
p <- ggplot(p_county, aes(x=reorder(County, Turnout), y=Turnout)) +
  geom_bar(stat="identity", width=.8, fill="white", color="black") +
  geom_text(aes(label=paste(Turnout, "%")), position=position_stack(vjust=.5), size=2.5) +
  labs(x="縣市（依投票率排序）", y="投票率",
       title="2020台灣總統大選各縣市：投票率") +
  coord_flip()

ggplotly(p, width=800, height=600, tooltip=NULL)
```

It is widely suggested that an increasing voter turnout due to younger generation this time can hugely impact the results.
Here we plot the turnout against DPP-KMT share differences by county:

```{r presidential_county_turnout_vsd, fig.showtext=TRUE, fig.cap="2020 Presidential Election Voter Turnout by County"}
ggplot(p_county, aes(x=Turnout, y=Share_Diff, label=County)) +
  scale_x_continuous(limits=c(35, 80)) +
  geom_rect(data=p_county[1],
            aes(xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf), fill="lightgreen", alpha=.8) +
  geom_rect(data=p_county[1],
            aes(xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=0), fill="lightblue", alpha=.8) +
  geom_hline(aes(yintercept=0), color="red") +
  geom_point() +
  geom_smooth(method="lm") +
  geom_text(check_overlap=TRUE, color="grey40") +
  labs(x="投票率％", y="綠藍得票率差",
       title="2020台灣總統大選各縣市：投票率Ｘ藍綠得票率差")
```

We zoom-in in the following plot for those with turnout > 70%:

```{r presidential_county_turnout70_vsd, fig.showtext=TRUE, fig.cap="2020 Presidential Election Voter Turnout (>70%) by County"}
ggplot(p_county[Turnout > 70], aes(x=Turnout, y=Share_Diff, label=County)) +
  scale_x_continuous(limits=c(70, 80)) +
  geom_rect(data=p_county[Turnout > 70][1],
            aes(xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf), fill="lightgreen", alpha=.8) +
  geom_rect(data=p_county[Turnout > 70][1],
            aes(xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=0), fill="lightblue", alpha=.8) +
  geom_hline(aes(yintercept=0), color="red") +
  geom_point() +
  geom_text_repel(color="grey40") +
  labs(x="投票率％", y="綠藍得票率差",
       title="2020台灣總統大選各縣市：投票率(>70%)Ｘ綠藍得票率差")
```

We can also visualize the turnout relation on each party candidate instead:

```{r presidential_county_turnout_vs, fig.showtext=TRUE, fig.width=8, fig.height=3, fig.cap="2020 Presidential Election Voter Turnout vs Candidate Vote Shares by County"}
ggplot(p_county_l, aes(x=Turnout, y=Votes_Share)) +
  geom_rect(data=unique(p_county_l, by="Candidate"), aes(fill=Candidate),
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf, alpha=.5) +
  scale_fill_manual(values=main_party_colors) +
  theme(legend.position="none") +
  geom_point() +
  geom_smooth(method="lm", color="black") +
  facet_grid(~ Candidate) +
  labs(x="投票率％", y="得票率",
       title="2020台灣總統大選各縣市：投票率Ｘ得票率")
```

[Go Back to Top](#head)

#### By County Sub-Region

```{r process_presidential_region_turnout}
# Compute region-level voting rate and vote shares in the long data.
p_region_l <- p_region_l[p_region, Turnout:=i.Turnout, on=.(Region)]
p_region_l <- p_region_l[, Votes_Share:=Votes / sum(Votes), by=.(Region)]
```

From the previous county-level visualization we realized that turnout may be related to the difference between DDP and KMT candidates.
Here we use a simple linear model to test the correlation based on region-level data:

```{r lm_turnout_vsd}
lm_turnout <- lm(Share_Diff ~ Turnout, data=p_region_l)
coeftest(lm_turnout, vcov=vcovHC(lm_turnout, type="HC0"))
```

Apparently the correlation is significant.
Let's also generate the region-level scatterplot:

```{r presidential_region_turnout_vsd, fig.showtext=TRUE, fig.cap="2020 Presidential Election Voter Turnout by Region"}
ggplot(p_region, aes(x=Turnout, y=Share_Diff)) +
  geom_rect(data=p_region[1], aes(xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf),
            fill="lightgreen", alpha=.8) +
  geom_rect(data=p_region[1], aes(xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=0),
            fill="lightblue", alpha=.8) +
  geom_hline(aes(yintercept=0), color="red") +
  geom_point() +
  geom_smooth(method="lm") +
  labs(x="投票率％", y="綠藍得票率差",
       title="2020台灣總統大選各鄉鎮市區：投票率Ｘ綠藍得票率差")
```

Again with additional zoom-in:

```{r presidential_region_turnout70_vsd, fig.showtext=TRUE, fig.cap="2020 Presidential Election Voter Turnout (>70%) by Region"}
ggplot(p_region[Turnout > 70], aes(x=Turnout, y=Share_Diff, label=Region)) +
  geom_rect(data=p_region[1], aes(xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf),
            fill="lightgreen", alpha=.8) +
  geom_rect(data=p_region[1], aes(xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=0),
            fill="lightblue", alpha=.8) +
  geom_hline(aes(yintercept=0), color="red") +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  geom_text(check_overlap=TRUE, nudge_y=-.035, color="darkgrey") +
  labs(x="投票率％", y="綠藍得票率差",
       title="2020台灣總統大選各鄉鎮市區：投票率(>70%)Ｘ綠藍得票率差")
```

We can also test the correlation between turnout and vote share of each party's candidates:

```{r lm_vr_vs}
for ( candidate in unique(p_region_l$Candidate) ) {
  lm_turnout_candidiate <- lm(Votes_Share ~ Turnout, data=p_region_l[Candidate == candidate])
  message(rep("-", 80))
  message("Test correlation between vote shares and voter turnout.")
  message(sprintf("For %s:", candidate))
  print(coeftest(lm_turnout_candidiate, vcov=vcovHC(lm_turnout_candidiate, type="HC0")))
}
```

Interestingly,
such correlation is not found for PFP candidates.

```{r presidential_region_vr_vs, fig.showtext=TRUE, fig.width=8, fig.height=3, fig.cap="2020 Presidential Election Voter Turnout vs Candidate Vote Shares by Region"}
ggplot(p_region_l, aes(x=Turnout, y=Votes_Share)) +
  geom_rect(data=unique(p_region_l, by="Candidate"), aes(fill=Candidate),
            xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf, alpha=.5) +
  scale_fill_manual(values=main_party_colors) +
  theme(legend.position="none") +
  geom_point() +
  geom_smooth(method="lm", color="black") +
  facet_grid(~ Candidate) +
  labs(x="投票率％", y="得票率",
       title="2020台灣總統大選各縣市：投票率 X 得票率")
```

[Go Back to Top](#head)

#### By Polling Place

TBC.

[Go Back to Top](#head)

## 第十屆立法委員選舉結果 The 10th Legislative Election Results {.tabset .tabset-fade .tabset-pills}

### 不分區立委 Party List {.tabset .tabset-fade .tabset-pills}

```{r process_party_list_results, results="hide"}
partylist <- fread("data/processed/legislative/legislative_第10屆全國不分區及僑居國外國民立法委員.csv", encoding="UTF-8")
partylist <- partylist[, Region:=trimws(Region, whitespace="[\\h\\v]")]  # Trim full-width white space.
cols_area <- c("By", "Region", "Village", "Place")
cols_party <- names(partylist)[7:25]  # There are 19 parties in total.
partylist_l <- melt(partylist[, c(cols_area, cols_party), with=FALSE],
                   id.vars=cols_area, variable.name="Party", value.name="Votes")
partylist_l <- partylist_l[, Party:=factor(Party)]
partylist_final <- partylist_l[, .(Votes=sum(Votes)), by=.(Party)]
partylist_final <- partylist_final[, Share:=Votes / sum(Votes)]
```

For party list to secure at least one seat in [Legislative Yuan](https://en.wikipedia.org/wiki/Legislative_Yuan),
it must obtain minimally 5% of party-list votes.
If a party cannot pass the 5% threshold but pass the 3% threshold,
it will be sponsored by the government in the following 4 years (based on number of votes obtained).
This is particularly important for small parties to cover their operating costs.

```{r partylist_final_result, fig.showtext=TRUE, fig.cap="2020 Legislative Election Party List Final Outcome"}
thres_003 <- sum(partylist_final$Votes) * .03
thres_005 <- sum(partylist_final$Votes) * .05

ggplot(partylist_final, aes(x=reorder(Party, Votes), y=Votes)) +
  geom_bar(stat="identity") +
  labs(x="政黨", y="得票數",
       title="2020台灣不分區立委選舉結果") +
  geom_text(aes(x=Party, y=Votes,
                label=sprintf("%s (%s)", scales::comma(Votes), scales::percent(Share, .01))),
            size=3, hjust=-.05) +
  scale_y_continuous(labels=scales::comma, limits=c(0, max(partylist_final$Votes) * 1.25)) +
  geom_hline(aes(yintercept=thres_003, color="補助 Sponsorship"), linetype=3) +
  geom_hline(aes(yintercept=thres_005, color="席次 Seats"), linetype=3) +
  scale_colour_manual(name="門檻 Threshold", values=c("red", "green")) +
  theme(legend.position="bottom") +
  coord_flip()
```

#### By County

We group parties with less than 2% votes into "others" to make the rest of the plots easier to read.

```{r process_party_list_county, results="hide"}
partylist_county <- partylist_l[, .(Votes=sum(Votes)), by=.(By, Party)]
small_parties <- partylist_final[Share < .02, Party]

partylist_county <- partylist_county[Party %in% small_parties, Party:="其他"]
partylist_county <- partylist_county[, Party:=factor(Party)]  # Renew level.
partylist_county <- partylist_county[, .(Votes=sum(Votes)), by=.(By, Party)]  # Re-group.
partylist_county <- partylist_county[, Share:=Votes / sum(Votes), by=.(By)]

# Compute DPP-KMT share differences.
partylist_county_gb <- partylist_county[Party %in% c("(14)  民主進步黨", "(9)  中國國民黨")]
setorder(partylist_county_gb, By, Party)
partylist_county_gb_diff <- partylist_county_gb[, .(Share_Diff=Share[2] - Share[1]), by=.(By)]
partylist_county_gb_diff <- partylist_county_gb_diff[, dpp_win:=Share_Diff > 0]
```

```{r party_list_county, fig.showtext=TRUE, fig.cap="2020 Legislative Election Party List Vote Shares by County"}
party_colors <- c("darkorange", "darkgreen", "yellow", "dodgerblue", "green", "green3", "lightblue", "grey")
ggplot(partylist_county, aes(x=reorder(By, Votes), y=Share, fill=Party)) +
  geom_bar(stat="identity", position="stack", width=.8) +
  geom_text(aes(label=scales::percent(Share, .01)), position=position_stack(vjust=.5), size=2, angle=20) +
  scale_fill_manual(name="政黨", values=party_colors) +
  labs(x="縣市（依總票數排序）", y="得票率",
       title="2020台灣不分區立委各縣市政黨得票率") +
  coord_flip() +
  theme(legend.position="bottom")
```

```{r party_list_dpp_kmt_diff, fig.cap="2020 Legislative Election Party List DPP-KMT Share Difference by County"}
p <- ggplot(partylist_county_gb_diff, aes(x=reorder(By, Share_Diff), fill=dpp_win, y=Share_Diff)) +
  geom_bar(stat="identity", width=.8, color="black", alpha=.5) +
  scale_fill_manual(values=c("dodgerblue", "green3")) +
  theme(legend.position="none") +
  geom_text(aes(label=scales::percent(Share_Diff)), position=position_stack(vjust=.5), size=2.5) +
  labs(x="縣市（依得票率差排序）", y="綠藍得票率差",
       title="2020台灣不分區立委各縣市：綠藍得票率差") +
  coord_flip()

ggplotly(p, width=800, height=600)
```

[Go Back to Top](#head)

#### By County Sub-Region

```{r process_party_list_region, results="hide"}
# Process region-level party list results.
partylist_region <- partylist_l[, .(Votes=sum(Votes)), by=.(By, Region, Party)]
partylist_region <- partylist_region[, Party:=factor(Party)]  # Renew level.
partylist_region <- partylist_region[, Share:=Votes / sum(Votes), by=.(By, Region)]
partylist_region <- partylist_region[!Party %in% small_parties]  # Ignore small parties.

# Get presidential invalid vote share and turnout.
partylist_region <- partylist_region[p_region, `:=`(
  P_Invalid_Votes_Share=i.Invalid_Votes_Share,
  P_Turnout=i.Turnout
  ), on=.(By, Region)]

# Get presidential vote results.
p_region_l <- p_region_l[, P_Share:=Votes / sum(Votes), by=.(By, Region)]
partylist_region <- merge(partylist_region, p_region_l[, .(By, Region, Candidate, P_Share, Share_Diff)], by=c("By", "Region"), allow.cartesian=TRUE)

# Helper function for scatter plot.
scatterplot <- function() {
  ggplot(partylist_region, aes(x=Share, y=P_Share, color=Candidate)) +
    geom_rect(data=unique(partylist_region, by=c("Party", "Candidate")), aes(fill=Party),
              xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf, alpha=.25) +
    scale_fill_manual(values=party_colors) +
    geom_point(size=.5) +
    geom_smooth(method="lm", aes(color=Candidate)) +
    scale_color_manual(values=main_party_colors) +
    labs(x="總統候選人得票率", y="不分區政黨得票率",
         title="2020總統候選人得票率Ｘ不分區政黨得票率") +
    facet_wrap(Party ~ Candidate, ncol=3, scales="free")
}
```

We can plot the correlation between presidential candidate vote shares and party vote shares at region level.
But we must be very careful in interpreting the relation since we don't know the direction of such relation, if any.
Strictly speaking,
we don't even know whether they are causally related at all.

```{r, fig.height=15, fig.showtext=TRUE, fig.cap="2020 Legislative Election Party List Shares v.s. Presidential Votes"}
scatterplot()
```

[Go Back to Top](#head)

### 分區立委 Constituency {.tabset .tabset-fade .tabset-pills}

TBC.

[Go Back to Top](#head)

# References on R Packages
