---
title: "Topic Modeling: Latent Dirichlet Allocation"
subtitle: ""
author:
- name: Kyle Chung
  affiliation:
date: "`r format(Sys.time(), '%d %b %Y')` Last Updated (23 July 2020 First Uploaded)"
output:
  html_notebook:
    highlight: tango
    number_sections: yes
    theme: paper
    toc: yes
    toc_depth: 3
    toc_float: yes
    includes:
      in_header: /tmp/meta_header.html
  code_download: true
bibliography: lda.bib
link-citations: yes
abstract: |
  A technical introduction to LDA, the most popular topic modeling approach to date, along with working examples using R.
---

```{r meta, include=FALSE}
meta_header_file <- file("/tmp/meta_header.html")

# Add open graph meta.
meta <- c(
  '<meta name="author" content="Kyle Chung">',
  '<meta property="og:title" content="Latent Dirichlet Allocation for Topic Modeling">',
  '<meta property="og:type" content="article">',
  '<meta property="og:url" content="https://everdark.github.io/k9/notebooks/stats/lda/lda.nb.html">',
  '<meta property="og:image" content="https://everdark.github.io/k9/assets/boot.png">',
  '<meta property="og:description" content="A data science notebook about LDA.">'
)
contents <- meta

# Add Github corner.
github_corner_svg <- "../../../assets/github_corner.html"
github_corner_conf <- list(github_link="https://github.com/everdark/k9/tree/master/notebooks/stats/bootstrap")
contents <- c(contents, stringr::str_interp(readLines(github_corner_svg), github_corner_conf))
writeLines(contents, meta_header_file)

close(meta_header_file)
```

# LDA: A Bayesian Approach

LDA (Latent Dirichlet Allocation) is probably the most popular approach to date for topic modeling.
The approach formulates the problem as a generative process,
where a document is characterized by a distribution of topics,
and a topic in turn is characterized by a distribution of words.
In such hierarchical framework we can then represent (and indeed generate) a document based on bag-of-words.

As long as we can fully specify bith distribution mentioned above,
we are able to build a probabilistic model that can generate the document-term matrix in the following steps:

1. Draw a topic mixing from a document-topic distribution that represents the latent topics of a document. (Something like `70% topic-1, 20% topic-2 10% topic-3` for a 3-topic setup.) 
2. For each word position of the document, repeat:
    a. draw a topic from the topic mixing, and hence derive the corresponding topic-word distribution
    b. draw a word from the resulting topic-word distribution

The prior for both the document-topic distribution and the topic-word distribution are postulated as Dirichlet distribution,
where the parameters are set such that a sparse distribution is preferred.

For a detailed discussion on Bayesian modeling,
readers can refer to the notebook: [Bayesian Modeling Explained](https://everdark.github.io/k9/notebooks/stats/bayesian/bayesian_modeling_explained.nb.html).

**Not a Typical NLU Task**

In our [technical blog](https://everdark.github.io/k9/) We didn't put this notebook under the category of `Natural Language Understanding`,
instead we put it under the `Statistics` category.
This is to emphasize that the model is a pure probabilistic model based on bag-of-words,
without any mechanism of semantic embeddings.
It is powerful partly thanks to this simplicity.

# Dirichlet Distribution

Dirichlet distribution is just the multivariate version of Beta distribution.


# A Working Demo

@spm


```{r}
library(sentencepiece)


```


@text2vec

```{r wip}
library(text2vec)

data("movie_review")
tokens = tolower(movie_review$review[1:4000])
tokens = word_tokenizer(tokens)
it = itoken(tokens, ids = movie_review$id[1:4000], progressbar = FALSE)
v = create_vocabulary(it)
v = prune_vocabulary(v, term_count_min = 10, doc_proportion_max = 0.2)
  
vectorizer = vocab_vectorizer(v)
dtm = create_dtm(it, vectorizer, type = "dgTMatrix")

lda_model = LDA$new(n_topics = 10, doc_topic_prior = 0.1, topic_word_prior = 0.01)
doc_topic_distr = 
  lda_model$fit_transform(x = dtm, n_iter = 1000, 
                          convergence_tol = 0.001, n_check_convergence = 25, 
                          progressbar = FALSE)
```

```{r}
barplot(doc_topic_distr[1, ], xlab = "topic", 
        ylab = "proportion", ylim = c(0, 1), 
        names.arg = 1:ncol(doc_topic_distr))
```


```{r}
lda_model$get_top_words(n = 10, topic_number = c(1L, 5L, 10L), lambda = 1)
```

```{r}
lda_model$get_top_words(n = 10, topic_number = c(1L, 5L, 10L), lambda = 0.2)
```

# Topic Interpretation

@sievert2014ldavis

## Term Relevancy

In their work the relevance of a term $w$ to a specific topic $k$ is measred by

$$
r(w, k \vert \lambda) = \lambda \log(\phi_{kw}) + (1 - \lambda)\log(\frac{\phi_{kw}}{p_w}),
$$

where $\phi_{kw}$ is the probability of term $w$ appears in topic $k$,
and $p_w$ is the marginal probability of term $w$ in the corpus (marginalized over topics),
$\lambda$ is a parameter controlling the weight of term probability within topic against its lift (as measured by $\frac{\phi_{kw}}{p_w}$).

In this setup,
setting $\lambda = 1$ effectively rank the relevancy only in terms of the word probability within topic.
So the most popular word within the topic will be the most relevant word as well.
If a word is popular over the entire corpus (a common word) it will likely to appear in higher rank as well,
which is not helping in interpreting the given topic.

In their study on determining the optimal $\lambda$ they use the [Newsgroup](http://qwone.com/~jason/20Newsgroups/) dataset with human testing and find that $\lambda = 0.6$ is the best choice,
at least for their particular model.

The lesson is that we should never choose a trivial $\lambda$ value on the boundary,
i.e., neither 1 nor 0,
but experiment with a value in-between.

## Topic Distance

In `LDAvis` the left panel visualization by default will show the top-2 principal components on topc distances,
measured by [Jensen-Shannon divergence](https://en.wikipedia.org/wiki/Jensen%E2%80%93Shannon_divergence).
It is a measure to compute the distance between two distributions.

Remember that in LDA a topic is modeled as a word distribution.
Given any two topics we'd like to compute their distributional distance to understand how well we cluster the documents into topic groups.

The Jensen-Shannon divergence is a symmetric version of the [Kullback–Leibler divergence](https://en.wikipedia.org/wiki/Kullback%E2%80%93Leibler_divergence),
or relative entropy.
The KL divergence of a distribution P from a distribution Q is defined as (assuming a discrete setup):

$$
D_{KL}(P \vert\vert Q) = \sum_xP(x)\log\frac{P(x)}{Q(x)}.
$$

It is the expectation of logarithmic difference between P and Q.

The Jensen-Shannon divergence is defined as:

$$
JSD(P, Q) = \frac{1}{2}D_{KL}(P\vert\vert M) + \frac{1}{2}D_{KL}(Q\vert\vert M),
$$

where

$$
M = \frac{1}{2}(P + Q),
$$

is the mixture of the two distribution.

If we expaned the above equation:

$$
\begin{aligned}
JSD(P, Q) 
&= \frac{1}{2}D_{KL}(P\vert\vert M) + \frac{1}{2}D_{KL}(Q\vert\vert M) \\
&= \frac{1}{2}\sum_xP(x)\log\frac{P(x)}{M(x)} + \frac{1}{2}\sum_xQ(x)\log\frac{Q(x)}{M(x)} \\
&= \frac{1}{2}\bigg[\sum_xP(x)\log P(x) + \sum_xQ(x)\log Q(x)\bigg] -
\sum_x\bigg(\frac{P(x) + Q(x)}{2}\bigg)\log\bigg(\frac{P(x) + Q(x)}{2}\bigg) \\
&= H_M - \frac{H_P + H_Q}{2},
\end{aligned}
$$

where $H_D$ is the [entropy](https://en.wikipedia.org/wiki/Entropy) of distribution $D$:

$$
H_D = - \sum_xD(x)\log D(x).
$$

Hence the JSD is indeed *the entropy of the mixture minus the mixture of the entropy.*

Once the JSD between all topics are computed,
[multidimensional scaling](https://en.wikipedia.org/wiki/Multidimensional_scaling) is used to convert the distance matrix into a 2-dimentional coordinates for visaulization purpose.

## Demo


```{r}
#lda_model$plot()
## TODO: how to show in notebook?
```


# References
